---
title: "Final_INF 6480"
author: "Leslie Garcia"
date: "2025-12-13"
output: html_document
---

Load packages 
```{r}
library(tidyverse)
library(lubridate)
library(broom)
library(reshape2)
library(scales)
library(dplyr)
```

Read data file 
```{r}
border <- read.csv("Border_Crossing_Entry_Data_072025.csv")
```

Filter to California ports data only
```{r}
border_ca <- border %>%
  filter(State == "California")

glimpse(border_ca)
nrow(border_ca) #should be > 0
table(border_ca[['State']])
```

Clean data and create holiday indicators  
```{r}
border_ca_clean <- border_ca %>%
mutate (
  Date = my(Date),
  Year = year(Date),
  Month = month(Date),
  Month_Name = month(Date, label = TRUE, abbr = TRUE),

# ORIGINAL holiday definition (July, Nov, Dec)
HolidayMonth = if_else(Month %in% c(7, 11, 12),
                       "Holiday", "Non-Holiday"),
HolidayMonth = factor(HolidayMonth),

# NEW: Spring/Easter months (March & April)
Spring_Easter_Month = Month %in% c(3, 4),

# NEW: Memorial Day month (May)
Memorial_Month = Month == 5,

# OPTIONAL: broader "major travel" months
# (Spring break/Easter, Memorial Day, July 4, Thanksgiving, Christmas/New Year)
Holiday_MajorTravel = if_else(Month %in% c(3, 4, 5, 7, 11, 12),
                              "Holiday", "Non-Holiday"),
Holiday_MajorTravel = factor(Holiday_MajorTravel),

Border = factor(Border),
Measure = factor(Measure),
`Port.Name` = factor(`Port.Name`),
Value_log = log1p(Value),
Lat_scaled = scale(Latitude)[,1], 
Lon_scaled = scale(Longitude)[,1]

) %>%
filter(!is.na(Date), !is.na(Value))

summary(border_ca_clean)

```

Check for months labeled as holiday
```{r}
border_ca_clean %>%
  count(Year, Month, Month_Name, HolidayMonth, Holiday_MajorTravel,
        Spring_Easter_Month, Memorial_Month) %>%
  arrange(Year, Month) %>%
  head(50)

```

-Check the counts
```{r}
table(border_ca_clean$HolidayMonth)
table(border_ca_clean$Holiday_MajorTravel)
```

-quick barplot to check labeling 
```{r}
table(border_ca_clean[["Holiday_MajorTravel"]])
```

#quick viz to check holiday labeling
```{r}
ggplot(border_ca_clean, aes(x = Month_Name, fill = HolidayMonth)) +
  geom_bar() +
  labs(title = "Count of Observations by Month (Holiday vs Non-Holiday)",
       x = "Month", y = "Number of Records")

```

Check major travel holiday labels

```{r}
ggplot(border_ca_clean, aes(x = Month_Name, fill = Holiday_MajorTravel)) +
geom_bar() +
labs(title = "Observations by Month (Major Travel Holiday Definition)",
x = "Month", y = "Number of Records")

```

Monthly totals 
```{r}
monthly_ca <- border_ca_clean %>%
group_by(Date, Month, HolidayMonth) %>%
summarise(total_value = sum(Value), .groups = "drop")

max_total <- max(monthly_ca[["total_value"]], na.rm = TRUE)

```


DESCRIPTIVE STATISTICS

```{r}
#overall summary 
desc_ca <- border_ca_clean %>%
  summarise(
    mean_value = mean(Value),
    median_value = median(Value),
    sd_value = sd(Value),
    var_value = var(Value)
  )
desc_ca

#Holiday/non-holiday
desc_ca_holiday <- border_ca_clean %>%
  group_by(HolidayMonth) %>%
  summarise(
    mean_value = mean(Value),
    median_value = median(Value),
    sd_value = sd(Value),
    n = n()
  )
desc_ca_holiday

#Frequency of transportation modes
table(border_ca_clean$Measure)
prop.table(table(border_ca_clean$Measure))

```


VISUALIZATION 
- # Histogram of crossing counts
```{r}
ggplot(border_ca_clean, aes(Value)) +
  geom_histogram(bins = 30) +
  labs(title = "Distribution of California Border Crossing Counts",
       x = "Crossing Count", y = "Frequency")
```

- Boxplot: Holiday vs Non-Holiday
```{r}
ggplot(border_ca_clean, aes(HolidayMonth, Value)) +
  geom_boxplot() +
  scale_y_continuous(labels = comma) +
  labs(title = "California Border Crossings: Holiday vs Non-Holiday",
       x = "Month Type", y = "Crossing Count")
```

- Time series trend
```{r}
monthly_ca <- border_ca_clean %>%
  group_by(Date) %>%
  summarise(total_value = sum(Value))

ggplot(monthly_ca, aes(Date, total_value)) +
  geom_line(size = 1) +
  geom_point(color = "blue", size = 1) +
  scale_y_continuous(labels = comma) +
  labs(title = "Monthly Border Crossings in California Over Time",
       x = "Date", y = "Total Crossings")
```

- Trend by holiday 
```{r}
monthly_ca_holiday <- border_ca_clean %>%
  group_by(Date, HolidayMonth) %>%
  summarise(total_value = sum(Value), .groups="drop")

ggplot(monthly_ca_holiday,
       aes(Date, total_value, color = HolidayMonth)) +
  geom_line() +
  labs(title = "California Monthly Border Crossings by Holiday Status",
       x = "Date", y = "Total Crossings")
```

- Correlation heat map for numeric variables
```{r}
num_vars_ca <- border_ca_clean %>%
  select(Value, Value_log, Latitude, Longitude)

cor_mat_ca <- cor(num_vars_ca, use = "complete.obs")
cor_melt_ca <- melt(cor_mat_ca)

ggplot(cor_melt_ca, aes(Var1, Var2, fill = value)) +
  geom_tile() +
  geom_text(aes(label = round(value, 2))) +
  labs(title = "Correlation Heatmap (California Data)",
       x = "", y = "")
```

HYPOTHESIS TESTING 
- T-test: Are crossings higher during holiday months?
```{r}
ttest_ca <- t.test(Value_log ~ HolidayMonth, data = border_ca_clean)
ttest_ca
```

- ANOVA: Differences by transportation mode
```{r}
anova_ca <- aov(Value_log ~ Measure, data = border_ca_clean)
summary(anova_ca)

##if significant add 
#TukeyHSD(anova_ca)

```



- Chi-square: Is HolidayMonth associated with Measure type?
```{r}
ca_tab <- table(border_ca_clean[["Measure"]], border_ca_clean[["HolidayMonth"]])
chisq.test(ca_tab)

```

- Correlation: Latitude vs crossing counts

```{r}
cor(border_ca_clean[["Value"]], border_ca_clean[["Latitude"]])
```

LINEAR REGRESION MODEL 
- Modeling dataset
```{r}
model_ca <- border_ca_clean %>%
  select(Value_log, HolidayMonth, Measure,
         Year, Month, Lat_scaled, Lon_scaled) %>%
  na.omit()

nrow(model_ca)
```

- Train-test split
```{r}
set.seed(123)
n <- nrow(model_ca)
train_idx <- sample(seq_len(n), size = 0.7*n)

train_ca <- model_ca[train_idx, ]
test_ca <- model_ca[-train_idx, ]
```

- Fit linear regression model

```{r}
lm_ca <- lm(
  Value_log ~ HolidayMonth + Measure +
    Year + Month + Lat_scaled + Lon_scaled,
  data = train_ca
)

summary(lm_ca)

```

- Predictions on test set
```{r}
test_ca[["pred_log"]] <- predict(lm_ca, newdata = test_ca)
head(test_ca)

```

- RMSE
```{r}
rmse_ca <- sqrt(mean((test_ca$pred_log - test_ca[["Value_log"]])^2))
rmse_ca

```

- Test R²
```{r}

sst <- sum((test_ca[["Value_log"]] - mean(test_ca[["Value_log"]]))^2)
sse <- sum((test_ca[["Value_log"]] - test_ca[["pred_log"]])^2)
r2_test_ca <- 1 - sse/sst
r2_test_ca

```

- Predicted vs Actual Plot
```{r}
ggplot(test_ca, aes(Value_log, pred_log)) +
  geom_point(alpha = 0.4) +
  geom_abline(slope=1, intercept=0, linetype="dashed") +
  labs(title = "Predicted vs Actual (California Border Crossings)",
       x = "Actual log(1 + Value)", y = "Predicted log(1 + Value)")
```

